---
title: "p8105_hw3_jt3649"
author: "Juan Tang"
date: "2025-10-09"
output: github_document
---

## Problem 1
```{r setup}
library(tidyverse)
library(patchwork)
#load the dataset
library(p8105.datasets)
data("instacart")
```

### Instacart dataset description
The instacart dataset contains `r nrow(instacart)` observations of `r ncol(instacart)` variables. Each row represents a single product from an order. The key variables include: `r paste(names(instacart))`. 

#### Describing variables
- `order_id` is order identifier. 
- `product_id` and `product_name` are variables for product identifiers. 
- `aisle_id` and `aisle` are aisle identifiers. 
- `department_id` and `department` are department identifiers. 
- `order_dow` is the day of the week that the order was placed. 
- `order_hour_of_day` is the hour of the day that the order was placed. 
- `days_since_prior_order` is the days since the last order. 

#### Illustrative example of observations
For example, the first observation shows that product "`r instacart$product_name[1]`" from the "`r instacart$aisle[1]`" aisle was ordered on day `r instacart$order_dow[1]` at hour `r instacart$order_hour_of_day[1]`.

### Answer questions
#### How many aisles are there, and which aisles are the most items ordered from?
```{r aisle_count}
aisle_count = instacart |>
  count(aisle, sort = TRUE) #group the rows with same aisle together and count how many rows fall into each group

n_aisle = nrow(aisle_count)
```
There are `r n_aisle` aisles in the dataset. The most items are ordered from: 
1. `r aisle_count$aisle[1]` (`r aisle_count$n[1]` items)
2. `r aisle_count$aisle[2]` (`r aisle_count$n[2]` items)
3. `r aisle_count$aisle[3]` (`r aisle_count$n[3]` items)

#### Make a plot that shows the number of items ordered in each aisle, limiting this to aisles with more than 10000 items ordered. 

```{r aisle_plot}
instacart |>
  count(aisle) |>
  #limit to aisle with more than 10000 items ordered
  filter(n > 10000) |> 
  mutate(aisle = fct_reorder(aisle, n)) |>
  #x-axis is aisle, y-axis is the count
  ggplot(aes(x = aisle, y = n, dpi = 300)) + 
  geom_col() + 
  #the plot axis label cannot be seen clearly, hence flip the xy axis for better view
  coord_flip() + 
  labs(
    title = "Number of Items Ordered by Aisle", 
    x = "Aisle", 
    y = "Number of Items Ordered"
  )

```

#### Make a table showing the three most popular items in each of the aisles “baking ingredients”, “dog food care”, and “packaged vegetables fruits”.

```{r popular_item}
instacart |>
  #include only aisle with character variable “baking ingredients”, “dog food care”, and “packaged vegetables fruits”
  filter(aisle %in% c("baking ingredients", "dog food care", "packaged vegetables fruits")) |>
  group_by(aisle, product_name) |>
  #count the number of orders per product
  summarize(n = n()) |>
  #sort by aisle and count in descending order
  arrange(aisle, desc(n)) |>
  #select only for the top 3 items per aisle
  slice_head(n = 3) |>
  knitr::kable(
    col.names = c("Aisle", "Product Name", "Number of Orders"), 
    caption = "Three Most Popular Items in Selected Aisle"
  )

```

#### Make a table showing the mean hour of the day at which Pink Lady Apples and Coffee Ice Cream are ordered on each day of the week
```{r mean_hour}
instacart |>
  filter(product_name %in% c("Pink Lady Apples", "Coffee Ice Cream")) |>
  # group by product and the day of week
  group_by(product_name, order_dow) |>
  # calculate the mean hour for each selected group
  summarize(mean_hour = mean(order_hour_of_day)) |>
  # reshape the data from long to wide
  pivot_wider(
    names_from = order_dow, # use the day of week as column names
    values_from = mean_hour # value in the table is the mean hour
  ) |>
  knitr::kable(
    digits = 2, # round to 2 decimal places
    col.names = c("Product", "Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"), 
    caption = "Mean Hour of Day for Orders on Each Day of the Week"
  )
```

### Problem 2
Import zillow dataset
```{r import_zillow_data}
rental_df = read_csv("dataset/zillow_data/Zip_zori_uc_sfrcondomfr_sm_month_NYC.csv") |>
  janitor::clean_names()
zip_code = read_csv("dataset/zillow_data/Zip Codes.csv") |>
  janitor::clean_names()
```

Clean and tidy Zipcode data
```{r clean_zipcode}
zip_code_clean = zip_code |>
  mutate(
    #add a borough column from county
    borough = case_when(
      str_detect(county, "New York") ~ "Manhattan", 
      str_detect(county, "Kings") ~ "Brooklyn", 
      str_detect(county, "Queens") ~ "Queens",
      str_detect(county, "Bronx") ~ "Bronx", 
      str_detect(county, "Richmond") ~ "Staten Island", 
      TRUE ~ NA_character_
    ), 
    #unify the county names in zip_code_tidy
    county = case_when(
      str_detect(county, "New York") ~ "New York", 
      str_detect(county, "Kings") ~ "Kings", 
      str_detect(county, "Queens") ~ "Queens",
      str_detect(county, "Bronx") ~ "Bronx", 
      str_detect(county, "Richmond") ~ "Richmond", 
      TRUE ~ county
      )
  )
```

Clean and tidy zillow rental data
```{r clean_rental_df}
rental_clean = rental_df |>
  #reshape from wide to long format
  pivot_longer(
    cols = starts_with("x"), #select col starting with "x"
    names_to = "date", 
    values_to = "price"
  ) |>
  mutate(
    #remove "x" prefix before the date
    date = str_remove(date, "x"), 
    #replace "_" with "-"
    date = str_replace_all(date, "_", "-"), 
    date = as.Date(date),
    year = year(date), 
    month = month(date)
  ) |>
  #filter out date not within 2015-01-01 and 2024-08-31 
  filter(date >= as.Date("2015-01-01"), date <= as.Date("2024-08-31")) |>
  rename(zip_code = region_name) |>
  #standardize the county names in rental_price
  mutate(
    county_name = str_remove(county_name, "County$"), 
    county_name = case_when(
      county_name == "New York" ~ "New York", 
      county_name == "Kings" ~ "Kings", 
      county_name == "Queens" ~ "Queens", 
      county_name == "Bronx" ~ "Bronx", 
      county_name == "Richmond" ~ "Richmond", 
      TRUE ~ county_name
      ) 
  ) |>
  #remove zip codes that have no price data
  filter(!is.na(price))
```


Merge rental and zipcode data
```{r zillow_merged}
zillow_merged = rental_clean |>
  left_join(zip_code_clean, by = "zip_code")
```

  
#### ZIP code observation frequency
```{r zip_code_frequency}
# count number of observations for each ZIP code
zip_count = rental_clean |>
  count(zip_code)
zip_count

# calculate the number of ZIP codes with complete data (116 observations)
n_116_obs = sum(zip_count$n == 116)

# calculate the number of ZIP codes with sparse data (<10 observations)
n_less_10_obs = sum(zip_count$n < 10)
```

`r n_116_obs` ZIP codes are observed 116 times. `r n_less_10_obs` ZIP codes are observed fewer than 10 times. 
Some ZIP codes are observed rarely because they may have been added to Zillow's coverage area more recently, or they may have missing data for certain months. It is also possible that areas with these ZIP codes have limited rental market data. 
The complete observations (116) indicate consistent rental market activity and data collection throughout the entire period (2015-01-01 to 2024-08-31).


#### Average rental price by borough and year
```{r borough_year_table}
zillow_merged |>
  # group by borough and year
  group_by(borough, year) |>
  # calculate the mean price, removed price = NA
  summarize(avg_price = mean(price, na.rm = TRUE)) |>
  pivot_wider (
    names_from = year, # year as col name
    values_from = avg_price
  ) |>
  knitr::kable(
    # round to integer
    digits = 0, 
    caption = "Average Rental Price by Borough and Year"
  )
```

##### Comment on the trend: 
The rental price have generally increased across all boroughs from 2015 to 2024. Manhattan consistently has the highest rental prices, followed by Brooklyn and Queens. Bronx and Staten Island have the most affordable rental prices. There is a growth in price for all five boroughs in 2021-2022. This can potentially due to the post-pandemic market changes. 

#### Plot showing NYC Rental Prices within ZIP codes for all available years. 
Your plot should facilitate comparisons across boroughs. Comment on any significant elements of this plot.
```{r rental_price_plot}
# create a line plot showing rental price changes over time by ZIP code
rental_plot = zillow_merged |>
  ggplot(aes(x = date, y = price, dpi = 300,
             group = borough, # group by borough
             color = borough) # use color to identify borough
         ) + 
  geom_line() + 
  # create separate panels for each borough
  facet_wrap(~borough, ncol = 2) + 
  labs(
    title = "NYC Rental Prices within ZIP Code Over Time", 
    subtitle = "2015-2024", 
    x = "Date", 
    y = "Rental Price ($)", 
    color = "Borough"
  ) 
rental_plot
```

##### Comment on the plot
The plot shows general upward trends in rental prices across all five boroughs. Manhattan has the largest difference in price levels each year and has the greatest fluctuation in rental prices over the years. Manhattan, Brooklyn, and Queens have a noticeable drop in rental prices around 2021, which might be related to the pandemic. But the drop is followed by recovery and growth. Price data for Staten Island is the most limited. 


#### Average rental price within each ZIP code over each month in 2023
```{r avg_monthly_rental_price}
dist_plot = zillow_merged |>
  filter(year == 2023) |> # keep only 2023 data
  # group by borough and zip_code
  group_by(borough, zip_code) |>
  # calculate avg price per ZIP code
  summarize(avg_price = mean(price, na.rm = TRUE)) |>
  ggplot(aes(x = borough, y = avg_price, fill = borough, dpi = 300)) + 
  geom_boxplot() + 
  labs(
    title = "Distribution of ZIP-Code-Level Rental Prices Across Boroughs (2023)", 
    x = "Borough", 
    y = "Average Rental Price ($)", 
    fill = "Borough"
  )
dist_plot
```

##### Comment on the plot
Manhattan has both the highest median rental price and the largest spread. There are also ZIP codes in Manhattan that appear as outliers, exceeding monthly rent of $5000. Brooklyn shows moderate prices with substantial variability. Queens, the Bronx, and Staten Island show lower median prices with less variability. The distribution highlights significant within-borough differences, especially in Manhattan and Brooklyn.


##### Combine and save the plots
```{r combined_plot}
combined_plot = rental_plot / dist_plot
ggsave("results/combined_ziLlow_plot.png", combined_plot, width = 12, height = 18)
```



### Problem 3
#### Import dataset
```{r import_data}
demo_df = read_csv("dataset/nhanes_covar.csv", skip = 4) |>
  janitor::clean_names()
accel_df = read_csv("dataset/nhanes_accel.csv") |>
  janitor::clean_names()
```

#### Clean, tidy, and merge the datasets
```{r clean_data}
demo_clean = demo_df |>
  drop_na() |> # remove rows with any missing values
  filter(age >= 21) |> # only include participants aged 21 or older
  mutate(
    sex = factor(sex, level = c(1, 2), labels = c("Male", "Female")), # convert sex from level to factor
    education = factor(education, levels = c(1, 2, 3), labels = c("Less than high school", "High school equivalent", "More than high school")), # convert education from levels to factor
  )

nhanes_merged = demo_clean |>
  left_join(accel_df, by = c("seqn"))
```

#### Number of men and women table
```{r demographic_table}
demo_table = demo_clean |>
  # group by education level and sex
  group_by(education, sex) |>
  # count the participants in each group
  summarize(n = n()) |>
  # use sex as col names and fill with number of men and women
  pivot_wider(
    names_from = sex, 
    values_from = n
  ) |>
  # draw the table
  knitr::kable(
    caption = "Number of Participants by Education and Sex"
  )
demo_table
```

#### Age distribution Plot
create a visualization of the age distributions for men and women in each education category. Comment on these items.
```{r demographic_plot}
demo_plot = demo_clean |>
  ggplot(aes(x = age, fill = sex, dpi = 300)) + 
           # create histogram with 5-year bins
           geom_histogram(binwidth = 5, position = "dodge") + 
           # create separate panels for each education level
           facet_wrap(~education) +
           labs(
             title = "Age Distribution by Sex and Education Level", 
             x = "Age", 
             y = "Number of Participants", 
             fill = "Sex"
           )
demo_plot
```

##### Comment on the plot
The distribution of participants is relatively balanced between males and females across education levels "Less than high school" and "High school equivalent". Those with more than high school education are younger on average and with more male participants, while those with less than high school education tend to be older. This reflects the generational differences in educational attainment. Overall, more than high school education has the most participants of both sexes.  

#### Total activity pattern
```{r total_activiy}
nhanes_activity = nhanes_merged |>
  # create new variable (sum)
  mutate(total_activity = rowSums(across(min1:min1440), na.rm = TRUE)
         ) |>
  select(seqn, sex, age, bmi, education, total_activity)
nhanes_activity
```

##### Total activity plot
```{r}
total_activity_plot = nhanes_activity |>
  ggplot(aes(x = age, y = total_activity, color = sex, dpi = 300)) + 
  geom_point(alpha = 0.5) + 
  geom_smooth() + # add trend line on scatterplot
  facet_wrap(~education) + # create separate panel for each education level
  labs(
    title = "Total Daily Activity by Age, Sex, and Education Level", 
    x = "Age (years)", 
    y = "Total Activity (mins)", 
    color = "Sex"
  )
total_activity_plot
```
##### Comment on plot
Total activity generally decreases with age across all education levels. The decline in activity with age for both men and women is most pronounced in those with less than high school education. There is a steep decline in activity after age 40 (for men) and age 60 (for women) in this group. Females tend to have slightly higher activity level than males in  "High school equivalent" and "More than high school" groups. 
In the more than high school group, the relative mild decline pattern suggests that higher education levels may be associated with more consistent and sustained physical activity across lifetime. 

#### Average 24-hour activity
```{r activity_pattern}
avg_activity_plot = nhanes_merged |>
  pivot_longer(
    cols = starts_with("min"), # select all minute columns
    names_to = "minute", # put col names in minute variable
    values_to = "mins", # put values of mins to "mins" variable
    names_prefix = "min" # remove "min" prefix
  ) |>
  mutate(minute = as.numeric(minute)) |> # convert minute from chr to numeric
  group_by(education, sex, minute) |>
  # calculate avg mins for each educational group
  summarize(mean_min = mean(mins, na.rm = TRUE)) |>
  ggplot(aes(x = minute, y = mean_min, color = sex, dpi = 300)) + 
  geom_line(alpha = 0.7) + 
  geom_smooth(se = FALSE, linewidth = 1.2) + # remove CI
  facet_wrap(~education) +
  labs(
    title = "24-Hour Activity Pattern by Education Level and Sex", 
    x = "Minute of Day", 
    y = "Average Mins", 
    color = "Sex"
  ) + 
  scale_x_continuous(
    breaks = c(0, 360, 720, 1080, 1440), # set break points
    labels = c("0:00", "6:00", "12:00", "18:00", "24:00") # add time labels
  )
avg_activity_plot
```

Saving plots
```{r saving_plots}
ggsave("results/age_distribution_plot.png", demo_plot)
ggsave("results/total_activity_plot.png", total_activity_plot)
ggsave("results/24h_activity_plot.png", avg_activity_plot)
```

##### Comment on plot
All groups show a clear pattern with low activity during nighttime hours (~0:00-6:00) and higher activity during waking hours (daytime hours). Activity peaks occur around midday to early afternoon. Participants with more than high school education show slightly higher activity levels and longer activity time during waking hours (~7:00-23:00) compared to other two groups. Sex differences are relatively modest, though females show slightly higher activity in the late morning to early afternoon periods across all education levels.
